---
title: "Class 07: Lasso Regression"
output: html_document
---

```{r, message=FALSE}
library(readr)
library(dplyr)
library(glmnet)
library(ggplot2)
theme_set(theme_minimal())
```

I want to start by constructing a data matrix with orthogonal columns. One
easy way to do this is by creating a random matrix, taking the SVD, and then
selecting just the left singular vectors (U). We know by construction that
the columns of U are orthogonal to one another:

```{r}
X <- svd(matrix(rnorm(100 * 4), ncol = 4))$u
```

Then, we will construct a response vector y:

```{r}
b <- c(10, 5, -3, 1)
y <- X %*% b + rnorm(100, sd=0.5)
t(X) %*% y # this is close, but not exactly the numbers given in the lab because of the noise term
```

## Running glmnet for lasso

To produce a lasso regression, we can just call the `glmnet` function with alpha=1 (this
is also the default and we could just not specify anything). Run glmnet with our X and y
values; do not specify the lambda or alpha terms (this produces a lasso regression along
a sequence of approximately 100 lamdba values).

```{r}

```

Now, run the code below to produce a sketch similar to the one that you did on the lab.
Does it match?!

```{r}
df <- tibble(beta = as.numeric(model$beta),
             lambda = rep(model$lambda, each=nrow(model$beta)),
             component = rep(sprintf("X%d", seq_len(4)), ncol(model$beta)))

ggplot(df, aes(lambda, beta)) + geom_line(aes(color = component))
```

Now, create a model object from `glmnet`, but specify alpha=0 to produce a ridge regression
fit:

```{r}

```

Run this code to see the model paths for ridge regression

```{r}
df <- tibble(beta = as.numeric(model$beta),
             lambda = rep(model$lambda, each=nrow(model$beta)),
             component = rep(sprintf("X%d", seq_len(4)), ncol(model$beta)))

df <- filter(df, lambda < 30) # adjust this to make the plot easier to view by
                              # restricting the scale of the x-axis
ggplot(df, aes(lambda, beta)) + geom_line(aes(color = component))
```

Unfortunately, we don't have an easy way to produce a best subset estimator using
glmnet. You will just have to live with your sketch!
