---
title: "Class 04: Linear Regression and Normal Equations"
output: html_document
---

### Setup

Start by loading a few R packages that we will need for today's class:

```{r}
library(readr)
library(ggplot2)
library(dplyr)
```

Next, let’s load in a small dataset to work with today. This is a well-known
statistics dataset that describes the fuel efficiency of various automobiles.

```{r}
cars <- read_csv("https://raw.githubusercontent.com/statsmaths/ml_data_ans/master/mpg.csv")
```

Today we are going to build linear regression models directly in R without
(at least at first) the use of the `lm` function.

### Making the model matrix

In the notes today, and really for almost the entire semester, we will be
starting with a matrix X that contains numeric variables that describe the
known features that we want to use to make predictions. You'll notice that the
data here, and most datasets you find in the wild, don't quite look like this
yet. To go from the `tibble` object that we have stored as `cars` to a matrix
object, we use the R function `model.matrix`. It is very flexible and provides
a lot of useful functions that we will see in upcoming weeks.

Specifically, the following code creates a model matrix using just the variables
`displ` (engine size) and cyl (the number of cylinders in the engine):

```{r}
X <- model.matrix(~ displ +  cyl , data = cars)
head(X)
```

Notice that R has added an intercept for us, which is quite helpful. The variable
we want to predict is fuel efficiency in city driving. This can be grabbed by
using the notation:

```{r}
y <- cars$cty
head(y)
```

We now have the components to compute the regression vector beta, but in
order to do this you'll need to know how to work with matrices in R.

### Matrix operations in R

Let's create two smaller matrices to work with:

```{r}
A <- matrix(c(1, 2, 3, 4, 5, 6), ncol = 3, byrow=TRUE)
A
```

And

```{r}
B <- matrix(c(1, 1, 2, 5), ncol = 2, byrow=TRUE)
B
```

To multiply two matrices in R, we use the symbol `%*%`:

```{r}
B %*% A
```

This also works if you have a vector and want to multiple it by a matrix:

```{r}
v <- c(4, 5, 6)
A %*% v
```

To take the transpose of a matrix, we use the operator `t`:

```{r}
t(B)
```

And to create an inverse, use the function `solve`:

```{r}
solve(B)
```

That's really all you should need to apply what we saw in class to the data
on the car fuel efficiency.

### Applying ordinary least squares

Compute the ordinary least squares solution from the data matrix and response
vector we created above (and store it as an object named `beta`) using the
matrix operations seen above:

```{r}

```

Now, produce a new object `yhat` that represents the predicted values from the
linear regression:

```{r}

```

Using the function `mean`, compute the mean squared error of your predictions
on the dataset:

```{r}

```

Use the `lm` function to verify that your `beta` matches the solution given
by R's function.

```{r}

```

### Indicator variables

Make a new model matrix that also includes the class (`class`) of the car:

```{r}

```

How does R convert the categorical variable `class` into numbers in the model
matrix? Construct the ordinary least square solution and compare the mean
squared error to the model that excludes class from the model:

```{r}

```

Does it do better with the new variable? Is it possible that the means squared
error could have gotten worse?
