---
title: "Class 02: Simple Linear Regression"
output: html_document
---

In addition to getting us started with regression, today will also serve as
an introduction to R and RStudio. I know some of you have used R, others
have used Python, some are CS majors who may have only used Java and C++,
and perhaps some of you have never written code before. All of these are
fine, but I want to get us all on the same page as quickly as possible. As
a first step, hopefully you have downloaded R and RStudio and are viewing
this *inside* R studio. If not, go back to the notes on the first day or
ask for help! If so, keep reading!

The document you are reading right now is written something called *Rmarkdown*.
The most powerful feature of Rmarkdown is that we can intermix code into the
document and actually run it in real time. To do this, we enclose code blocks
with three back ticks, and preface the first one with r in squiggly brackets.
Every code block will get run in sequence if when we hit the Preview or Knit
buttons, but we can also run just this block by hitting the play button to
the right of the block. Here, let’s add 1 and 1 together:

```{r}
1 + 1
```

This is very similar to IPython Notebooks, though some of the details are a
bit different. As a positive, the file is plain text and can be read and
edited easily in any generic text editor that you want to use. On the negative
side (depending on your view) is that the notebook does not store any of the
information from the executed code. You will always have to rerun the script
to get the output to work (this is only partially true; I can explain more if
you would like!).

### Install R Packages

In addition to the basic R functions that exist on start-up, there are
thousands of user-contributed packages the implement various add-ons. To
install these packages, we use the function install.packages. Run this code
to install the following packages; I wrote the code in such a way that it
it will not reinstall packages that you already have on your machine.

```{r}
if (!require("dplyr")) install.packages("dplyr")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("viridis")) install.packages("viridis")
if (!require("readr")) install.packages("readr")
if (!require("devtools")) install.packages("devtools")
```

Once a package is installed, we also need to load it. While installing the
package only needs to be done once, we have to load it each and every time we
restart R (notice up above that I included the option eval = FALSE in the code
block so that my computer does not constantly reinstall it):

```{r}
library(readr)
```

Once loaded, we can run commands from the readr package. We will do this in
just a moment.

### Loading data

Let’s load in four libraries that will be important for us throughout the
semester. Note that I’ll turn off messages as the packages produce quite a
bit of verbose output that we do not need to worry about. Also note that
re-loading the readr package has no ill-effects.

```{r}
library(readr)
library(ggplot2)
library(viridis)
library(dplyr)
```

Next, let’s load in a small dataset to work with today. This data consists of
the average number of hours per day that various species are awake. We will
read the data set, like most of those in this class, directly from my website:

```{r}
msleep <- read_csv("https://statsmaths.github.io/ml_data/msleep.csv")
```

Once this code has been run, you should see the dataset pop up in the upper
right hand corner of the screen. Clicking on it opens up a spreedsheet-like
view of the data.

Particularly important are the first three columns, as most datasets that we
work with this semester will have these columns as well. The meaning of these
are:

- the first column is just an id; don’t lose it!
- the second indicates whether this is a sample where you know the response
  or if have to predict the response
- the third column gives the response of interest; it is missing whenever the
  second column is equal to “test”

### qplot

We will use the function `qplot` to produce simple plots
based on our data. In its most basic form, we give `qplot`
the name of the variable we want to plot and the name of
the dataset in which the variable resides.

```{r, message = FALSE}
qplot(bodywt_log, data = msleep)
```

The function is quite smart and will choose by default the
most appropriate plot to use. Here it is creating a *histogram*
because we gave it one numeric variable.

Applying `qplot` to a categorical variable gives a bar plot:

```{r}
qplot(vore, data = msleep)
```

We can also give `qplot` two variable. For instance, if we
use two continuous variables the function will construct a
scatter plot:

```{r, message = FALSE}
qplot(bodywt_log, awake, data = msleep)
```

The missing values that we are warned about are those points
where we need to produce predictions.

### Linear regression

It seems like there is a positive relationship between body
weight and hours spent awake. Let's now try to model this using
a statistical learning model.

**NOW, PLEASE RETURN TO THE LAB QUESTIONS BEFORE PROCEEDING**

### Linear regression in R

We can add a linear regresion line to out plot using **qplot** directly.
Adding things to a `qplot` graphic literally involves using the `+` sign and
adding the results of other functions.
Here we use `geom_smooth` with an option to make the smooth strictly linear:

```{r}
qplot(bodywt_log, awake, data = msleep) +
  geom_smooth(method = "lm")
```

We can fit the exact some model analytically, rather than graphically,
by calling the `lm` function. Here we use an R formula: the
response variable (thing we want to predict), followed by the `~`
sign, followed by the predictor variable. As with the graphics command,
we need to indicate which dataset is being used. The output shows
the slope and intercept of the best-fit-line.

```{r}
model <- lm(awake ~ bodywt_log, data = msleep)
model
```

The function `predict` will give us the predicted values implied
by this line:

```{r}
predict(model, newdata = msleep)
```

### Finishing up

Finally, we want to test how well this linear regression performs on the
dataset. We can add the predictions back into the `msleep` dataset as follows:

```{r}
msleep$awake_pred <- predict(model, msleep)
```

Click on the msleep dataset again to verify that there is a
new column of predictions. Notice that these are filled in
for all values. We can check the mean squared error of these predictions
with the following code:

```{r}
tapply((msleep$awake_pred - msleep$awake)^2, msleep$train_id, mean)
```

Notice that, since you do not have the testing data, you cannot test how well
your prediction works on new data. Later, I will make this data public and
you can test your predictions. More on this in the upcoming weeks!





