---
title: "Exam 02: Take home exam"
output: html_document
---

```{r, message=FALSE}
library(readr)
library(dplyr)
library(glmnet)
library(ggplot2)
library(stringi)
theme_set(theme_minimal())
```

### Elastic net regression

Here, we will generate some random data with five columns. I included some
code to make sure that the columns of X are slightly correlated, otherwise
we would just get something that looked nearly the same as the case where
XtX is equal to the identity.

```{r}
set.seed(1)
n <- 100
p <- 5
X <- matrix(rnorm(n * p), ncol = p)

# make the columns of X moderately correlated with one another
X[,2] <- X[,2] + X[,1] * 0.5
X[,3] <- X[,3] + X[,1] * 0.5
X[,4] <- X[,3] + X[,1] * 0.5

# construct simulated values y
b0 <- c(2, -2, -0.5, 0,  0)
y <- X %*% b0 + rnorm(n, sd=0.5)
```

The function below applies the coordinate descent algorithm regression. You
will need to fill in the two lines where indicated (right now it just sets
beta[k] to 0 to) in order to implement the elastic net regression for a given
`lambda` and `alpha`.

```{r}
run_coord_descent_lasso <- function(X, y, lambda=1, alpha=1, max_iter=100)
{
  beta <- rep(0, ncol(X))                 # initialize beta with all zeros

  for (i in seq_len(max_iter))            # run coordinate descent for at most max_iter runs
  {
    beta_old <- beta                      # store previous beta value
    for (k in seq_along(beta))            # cycle through the coordinates
    {
      beta[k] <- 0                        # initialize beta_k to zero
      r <- y - X %*% beta                 # this is partial residual, because beta_k = 0
      rtx <- t(r) %*% X[,k]               # correlation of X_k and partial residual

      if (rtx > lambda)
      {
        # Condition to update beta[k] if rtx is larger than lambda
        beta[k] <- 0 # YOU NEED TO CHANGE THIS
      } else if (rtx < -1 * lambda) {
        # Condition to update beta[k] if rtx is less than -lambda
        beta[k] <- 0 # YOU NEED TO ALSO CHANGE THIS
      }
    }

    # check if anything changed; if not exit loop and return the result
    if (sum((beta - beta_old)^2) < 1e-10) { break }
  }

  return(beta)
}
```

Now, run coordinate descent on the dataset for these values of lambda (70)
and alpha (0.7):

```{r}
run_coord_descent_lasso(X, y, lambda=70, alpha=0.7)
```

Knit the markdown file to HTML, print out the result, and hand this in along
with your derivation of the elastic net coordinate descent terms.



