---
title: "Lab 14"
author: ""
output: html_notebook
---

# Set-up

Read in the following libraries and to load the metadata about
the flowers:

```{r}
library(readr)
library(dplyr)
library(glmnet)

f17 <- read_csv("https://statsmaths.github.io/ml_data/flowers_17.csv")
```

To get the actual data for the images, you'll have to download the following
file.

- https://github.com/statsmaths/ml_data/releases/download/v1/flowers_17_x64.rds

Once this is downloaded, you'll have to run something like this:

```{r}
x64 <- read_rds("~/Desktop/flowers_17_x64.rds")
```

If, for example, the file sits on your Desktop and you have a Mac. If you have
trouble with any of this, please let me know as soon as possible.


# Lab 14











# Submission

The code below assumes that you have added a prediction named
`pred` to every row of the dataset.

```{r}
submit <- select(flights, obs_id, train_id)
submit <- mutate(submit, pred = pred)
write_csv(submit, "class14_submit.csv")
```

Now, upload the csv file to GitHub and you are done.
