---
title: "Lab 15"
author: ""
output: html_notebook
---

# Set up

Read in the following libraries and to load the crimes dataset:

```{r}
library(readr)
library(dplyr)
library(ggplot2)
library(viridis)

crimes <- read_csv("https://statsmaths.github.io/ml_data/chi_crimes_12.csv")
```

The names of the crimes are, in order:

```{r}
crimes_list <- c("theft", "battery", "criminal damage", "narcotics",
"other offense", "assault", "burglary", "motor vehicle theft", "robbery",
"deceptive practice", "criminal trespass", "prostitution")
```

Make sure your predictions are all an integer from 1 to 12.

```{r}
if (!require("keras")) {
  install.packages("keras")
}
```

## Build neural networks






## Submitting your solutions

Finally, once you have your predictions saved as a variable called `pred`,
run the following code to produce your your results:

```{r}
submit <- select(flights, obs_id, train_id)
submit <- mutate(submit, pred = pred)
write_csv(submit, "class15_submit.csv")
```

Now, upload the csv file to GitHub and you are done.

