---
title: "Lab 20"
author: ""
output: html_document
---

# Set-up

Read in the following libraries and to load the metadata about
the flowers:

```{r}
library(readr)
library(dplyr)
library(glmnet)

stl <- read_csv("https://statsmaths.github.io/ml_data/stl_10.csv")
```

To get the embeddings, download this file as well:

```{r}
X <- as.matrix(read_csv("https://statsmaths.github.io/ml_data/stl_10_vgg16_cnn.csv"))
```




# Todays lab

The only thing you need to do today is to play around with different models
and see how well they work. Once you are satisfied, pick the best model and
incorporate the predictions into your output in the final chunk below. Note:
I would suggest keeping each model in its own code chunk for readability.
Also, save your work frequently as it is quite likely that you will crash
R at some point.










# Submission

The code below assumes that you have added a prediction named
`pred` to every row of the dataset.

```{r}
submit <- select(stl, obs_id, train_id)
submit <- mutate(submit, pred = pred)
write_csv(submit, "class20_submit.csv")
```

Now, upload the csv file to GitHub and you are done.
